Typed Functional Genetic Programming

### 2013-07-22 tex/thesis.tex
[tex]

@@

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem} 
\newtheorem*{proposition}{Proposition} 
\newtheorem{lemma}{Lemma} 
\newtheorem*{corollary}{Corollary}

@@

& ujistit se že je v části o tree reprezentacích 
  je napsaný co to znamená že podstrom má nějakej typ,
  že podstromy odpovídaj podvýrazum atd

 & říct že to je přesně ta notace co používá koza
 & EXAMPLES of tree representations of \lterms  
 & napsat transformace mezi \atTree a \sexprTree
 & nějak to sjednotit s tim uplně na začátku v popisu GP kde se popisuje
   přímo \sexprTree a říct proč se tomu tak řiká 
   (na tom místě co je v textu dřív)

 & zkontrolovat že reference tree-reps se odkazuje na dobrý místo,
   mohlo by se to rozbít tim že tree reprezentations - typovany/netypovany 
   jsou od sebe oddelený
\end{todo}

@@

\begin{todo}
   & říct že se vždycky zastavěj (strong normalization),
     && potřebný definice pro to uvést v \lterms sekci
     && musí se říc t co, tzn simpli typed l. kalkulus termy
        a betaeta-redukce
     && řekl bych že eta redukce je SN vždycky (tzn i untyped lk)
        pač plati (CR+terminující pak SN) a je CR vzhledem k barendrecht-stará-žlutá
        a terminující je rpotože zmenšuje term s nemuže zmenšovat donekonečna    
   & možná zmínit že se jedná o \textit{gentzen style proof}
   & říct že jsou to modus ponens a jeden směr věty o dedukci  
\end{todo}

\subsection{Tree representations of typed \lterms}

@@

More specifically,
\lets consider \textit{\sexprTree} $t$ corresponding to a \lterm of the type
$\sigma$ with root $r$ and subtrees $s_1 , \dots , s_n$. 
Then corresponding \textit{typed \sexprTree} $TT_S[t]$ for typed \lterm is 
obtained from the tree $t$ as follows.

\begin{equation*}
\mbox{ 
TT[
\Tree
	[.r 	
	  	  \text{$s_1$}
		  \text{$s_2$}
		  \text{$\dots$}
		  \text{$s_n$}
	] 
}]=
\mbox{
\Tree
	[.\text{$\sigma$ }
	    [.r 	
	  	  \text{$TT[s_1]$}
		  \text{$TT[s_2]$}
		  \text{$\dots$}
		  \text{$TT[s_n]$}
		]	  	
	] 
}
\end{equation*}

@@

@@

\begin{todo}
 & co to znamená mít lnf
 & potřebuju eště nějaký \textbf{propositions} z tyhle knihy?
 & ta definice je tam daná pro $\tur{}{M}{\sigma}$,
   ukazat ze ta naše je ok, že tam má navíc to $\Gamma$.
   Stačí udělat tríček že beru funkci co má lambda hlavu odpovídající 
   tomu kontextu.
\end{todo}

@@

The second rule can be replaced by more effective one.
\[ 
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
\]

@@

\begin{proposition}
Let context $\Gamma$, \lterm $M$ and type $\sigma$ be given.\\
Let $\garr$ be transitive reflexive closure of $\gar$. 
Then
$$ 
(\sigma,\Gamma) \garr M 
\Leftrightarrow
\tur{\Gamma}{M}{\sigma} \text{ and $M$ is in lnf}.   
$$
\end{proposition}

@@

\begin{todo}
 & zmínit že podobně jako mužeme transformovat inference rules
   na grammar rules, mužeme i obráceně grammar rules
   převést na inference rules - to bylo ostatně uděláno v tom dukazu ale
   je to podle mě hezký říct ty pravidla explicitně.
\end{todo}

@@

	\item By \textit{solving and-or tree} $T$ is meant finding finite tree $T'$
	      subgraph of $T$ such that following conditions are met. 
	    \begin{enumerate}
		  \item The root of $T'$ is the root of $T$.
		  \item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
		  \item Each \textit{or-node}  in $T'$ has precisely one child 
		      node.\footnote{\textbf{TODO}: MENTION why precisely one 
		      and not at least one ..or CHANGE the def. }   
	    \end{enumerate}

@@

The motivation for defining \textit{solving} of a And-or tree the way we did is that
a found tree $T'$ corresponds to generated \lterm. 
In order to understand this correspondence let's now talk about various tree representations
of \lterms.

\begin{todo}
 & Přepsat poslední odstavec --- počítal ještě s tím že 
   pasaže o stromové reprezentaci jsou až po této kapitole, což bylo změněno.
\end{todo}

\section{ Generating terms in Long normal form by solving Inhabitation tree }

Now we can finally put the pieces together. 
Every solution to an Inhabitation tree has this just described tree form of a typed \lterm.

\begin{todo}
& !!!
& TALK (more?) ABOUT "Barendregt-like" subsection \ref{barlike}\\
		Things about $\beta\eta^{-1}$ normal form, etc.   
\end{todo}

@@

\begin{todo}
 & několik fází, několik reprezentací
 & během generování se reprezentace mění 
   && nejdří obsahuje lambdy i volný proměnný pak už ne
      &&& vygeneruje se v \sexprTree
      &&& ta se převedena na \atTree aby jsme provedli abstraction elimination
 & dvě možnosti jak uchovávat jedince po vygenerování
   && \sexprTree
      &&& currently implemented
   && \atTree
      &&& currently not implemented, ale neni problém dodělat
      &&& DOĎELAT! :)
   && ale de mezi nima volně převádět
      &&& díky tomu že nemaj lambda abstrakce, jinak by museli bejt ve 
          WHNF/HNF nebo něco takovýho
   && dusledky pro křížení
      &&& \atTree umožňuje víc křížení než \sexprTree
          &&&& každý co de v \sexprTree umí i \atTree ale obráceně to neplatí
      &&& \atTree je zase ekvivalentní tomu kozovskýmu  
 & říct že bilding blocks je prostě kontext
\end{todo}

@@

In order to describe our state space to search in \lets introduce
extended version of our tree representation for typed \lterms from \ref{tree-reps}.
This extension consists in adding a new kind of leaf node $(\sigma,\Gamma)$ 
standing for unfinished tree of a type $\sigma$ with context $\Gamma$ to use
as set of building blocks
\footnote{\textbf{TODO}: Tady to napsat nějak líp s tim gamma }.\\

@@

\begin{todo}
& eště se musí nějak prohrotit to, že tam je navíc ta lambda hlava
kerá ještě o něco kazí tu $depth$
& ukázat že je to fakt to samý co kozovský klasický
& zavýst tam pojem běhu pro ten jeden vnější while cyklus
\end{todo}

@@

>>>>

\begin{todo}
 & Dozjistit a dopsat jak to Koza má s vyžadovánim unikátnosti při generování.
\end{todo}

@@

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem} 
\newtheorem{proposition}{Proposition} 
\newtheorem{lemma}{Lemma} 
\newtheorem*{corollary}{Corollary}

@@

\begin{todo}
 & napsat transformace mezi \atTree a \sexprTree
%& ujistit se že je v části o tree reprezentacích 
%  je napsaný co to znamená že podstrom má nějakej typ,
%  že podstromy odpovídaj podvýrazum atd
% & EXAMPLES of tree representations of \lterms  
% & nějak to sjednotit s tim uplně na začátku v popisu GP kde se popisuje
%   přímo \sexprTree a říct proč se tomu tak řiká 
%   (na tom místě co je v textu dřív) 
% & zkontrolovat že reference tree-reps se odkazuje na dobrý místo,
%   mohlo by se to rozbít tim že tree reprezentations - typovany/netypovany 
%   jsou od sebe oddelený
\end{todo}

@@

\begin{todo}
   & říct že se vždycky zastavěj (strong normalization)
     && a jaký to má výhody pro nás v GP (duležitý),
     && potřebný definice pro to uvést v \lterms sekci
     && musí se říct co je strongly normalizing, 
        tzn simpli typed $lambda$-kalkulus termy a $\beta$-redukce /
        $\beta\eta$-redukce
%     && řekl bych že eta redukce je SN vždycky (tzn i untyped lk)
%        pač plati (CR+terminující pak SN) a je CR vzhledem k barendrecht-stará-žlutá
%        a terminující je rpotože zmenšuje term s nemuže zmenšovat donekonečna    
   & možná zmínit že se jedná o \textit{gentzen style proof}
   & říct že jsou to modus ponens a jeden směr věty o dedukci  
\end{todo}

\subsection{Tree representations of typed \lterms}
\label{typed-tree-reps}

@@

More specifically,
\lets consider \textit{\sexprTree} $t$ corresponding to a \lterm of the type
$\sigma$ with root $r$ and subtrees $s_1 , \dots , s_n$. 
Then corresponding \textit{typed \sexprTree} $TT[t]$ for typed \lterm is 
obtained from the tree $t$ as follows.

\begin{equation*}
\mbox{ 
TT[ 
\Tree
	[.r 	
	  	  \text{$s_1$}
		  \text{$s_2$}
		  \text{$\dots$}
		  \text{$s_n$}
	] 
]} =
\mbox{
\Tree
	[.\text{$\sigma$ }
	    [.r 	
	  	  \text{$TT[s_1]$}
		  \text{$TT[s_2]$}
		  \text{$\dots$}
		  \text{$TT[s_n]$}
		]	  	
	] 
}
\end{equation*}

@@

%And analogically we can define transformation $TTtoTerm[]$ as follows.

@@

\begin{todo}
 & potřebuju eště nějaký \textbf{propositions} z \cite{barendregt10} tu uvíst ?
 & ta definice je tam daná pro $\tur{}{M}{\sigma}$,
   ukazat ze ta naše je ok, že tam má navíc to $\Gamma$.
   Stačí udělat tríček že beru funkci co má lambda hlavu odpovídající 
   tomu kontextu.
\end{todo}

@@

The second rule can be replaced by more effective one.
\[ 
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
~~~~ \textbf{if } n > 0
\]

@@

\begin{proposition}
\label{gram-lnf-prop}
Let context $\Gamma$, \lterm $M$ and type $\sigma$ be given.\\
Let $\garr$ be transitive reflexive closure of $\gar$. 
Then
$$ 
(\sigma,\Gamma) \garr M 
\Leftrightarrow
\tur{\Gamma}{M}{\sigma} \text{ and $M$ is in lnf}.   
$$
\end{proposition}

@@

\begin{todo}
 & možná zmínit že podobně jako mužeme transformovat inference rules
   na grammar rules, mužeme i obráceně grammar rules
   převést na inference rules - to bylo ostatně uděláno v tom dukazu ale
   je to podle mě hezký říct ty pravidla explicitně.
\end{todo}

@@

%	\item By \textit{solving and-or tree} $T$ is meant finding finite tree $T'$
%	      subgraph of $T$ such that following conditions are met. 
    \item $T'$ \textit{solves} and-or tree $T$ if $T'$ satisfy following conditions.
	    \begin{enumerate}
	      \item $T'$ is a finite tree, subgraph of $T$. 
		  \item The root of $T'$ is the root of $T$.
		  \item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
		  \item Each \textit{or-node}  in $T'$ has precisely one child 
		      node.\footnote{\textbf{TODO}: MENTION why precisely one 
		      and not at least one ..or CHANGE the def. }   
	    \end{enumerate}

@@

\newpage
\section{ Generating terms in \lnf by solving inhabitation tree }

The motivation for defining \textit{solving} and-or tree the way we did is that
solution trees of $\IT(\sigma,\Gamma)$ are  
\textit{typed \sexprTree{}s} representing all \lterms M in \lnf such that \GMS.

\begin{proposition}
\label{gram-it-prop}
(Correctness)\\
If \sexprTree $M'$ solves $\IT(\sigma,\Gamma)$,\\ 
then $M'$ represents \lterm $M$ in \lnf such that \GMS. 
\end{proposition}
\begin{proof}

We will prove following statement from which follows 
the rest by proposition \ref{gram-lnf-prop}:
If \sexprTree $M'$ solves $\IT(\sigma,\Gamma)$,
then $M'$ represents \lterm $M$ such that $(\sigma,\Gamma) \garr M$.\\

By induction on size of $M'$.\\

\textbf{If} $\sigma = \alpha$ such that $\alpha \in A$, then

\mbox{
$M' = $ 
\Tree 
[.$\alpha$ 
   [.$f$ $M'_1$ $\dots$ $M'_m$ ]
]}\\

For some $f : \rho_1 \ar \dots \ar \rho_m \ar \alpha \in \Gamma$.\\

One can easily see that $M'_i$ solves $\IT(\rho_i,\Gamma)$.

Thus by induction hypothesis we get that

$M'_i$ represents \lterm $M_i$ such that $(\rho_i,\Gamma) \garr M_i$.\\

$(\sigma,\Gamma) \gar (~f~(\rho_1,\Gamma)~\dots~(\rho_m,\Gamma)~)$,
thus $(\sigma,\Gamma) \garr (~f~M_i~\dots~M_i~)$.

Since $M'$ represents $(~f~M_i~\dots~M_i~)$, the proof for the first case is complete.\\

\textbf{Otherwise}, for $\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha$ 
such that $n> 0, \alpha \in A$

\mbox{$M' = $
\Tree [.$\tau_1\ar\dots\ar\tau_n\ar\alpha$ 
 [.$\lh{x_1~\dots~x_n}$ [.$N'$ ] ] ]    }

Again, one can easily see that $N'$ solves 
$\IT(\alpha;\Gamma,x_1:\tau_1,\dots,x_n:\tau_n)$.

Thus by induction hypothesis we get that

$N'$ represents \lterm $N$ such that 
$( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  ) \garr N$.\\

$
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
$,

thus $(\sigma,\Gamma) \garr (~\lambda~x_1~\dots~x_n~.~N~)$.

Since $M'$ represents $(~\lambda~x_1~\dots~x_n~.~N~)$, the proof is complete.

\newpage
\begin{proposition}
(Completeness)\\
If $M$ is \lterm in \lnf such that \GMS,\\
then there exists exactly one \sexprTree $M'$ such that 
$M'$ represents $M$ and
$M'$ solves $\IT(\sigma,\Gamma)$.
\end{proposition}
\begin{proof}
We will prove following statement from which follows 
the rest by proposition \ref{gram-lnf-prop}:
If $M$ is \lterm such that $(\sigma,\Gamma) \garr M$,
then there exists exactly one \sexprTree $M'$ such that 
$M'$ represents $M$ and
$M'$ solves $\IT(\sigma,\Gamma)$.\\

By induction on the length of rewriting chain of $(\sigma,\Gamma) \garr M$.\\

\textbf{If} $\sigma = \alpha$ such that $\alpha \in A$, then:

$(\sigma,\Gamma)$ must rewrite to
$(~f~( \rho_1 , \Gamma )~\dots~( \rho_m , \Gamma )~)$.

And there is only one possibility for $f$ since $M$ is given.

$M = (~f~M_1~\dots~M_m~)$ where $( \rho_i , \Gamma ) \garr M_i$.\\

From induction hypothesis we have that 
there exists exactly one \sexprTree $M_i'$ such that 
$M_i'$ represents $M_i$ and
$M_i'$ solves $\IT(\rho_i,\Gamma)$.\\

In order to construct $M'$ solving $\IT(\sigma,\Gamma)$ 
we have only one possibility since we know the right $f$ to take:

\mbox{
$M' = $ 
\Tree 
[.$\alpha$ 
   [.$f$ $M'_1$ $\dots$ $M'_m$ ]
]}\\

\textbf{Otherwise}, for $\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha,$ $(n > 0)$:\\

$(\sigma,\Gamma)$ must rewrite to
$(~\lambda~x_1~\dots~x_n~.~( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n )~)$.

$M = (~\lambda~x_1~\dots~x_n~.~N~)$ where 
$( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n ) \garr N$.\\

From induction hypothesis we have that 
there exists exactly one \sexprTree $N'$ such that 
$N'$ represents $N$ and
$N'$ solves $\IT(\alpha,\Gp)$.\\

In order to construct $M'$ solving $\IT(\sigma,\Gamma)$ in non-atomic $\sigma$
we have only one possibility:

\mbox{$M' = $
\Tree [.$\tau_1\ar\dots\ar\tau_n\ar\alpha$ 
 [.$\lh{x_1~\dots~x_n}$ [.$N'$ ] ] ]    }

\begin{todo}
& skontrolovat ty dukazy pořádně sepsal sem 
  je dost rychle a mužou bejt divný. 
& A možná to ještě to nějak okomentovat
\end{todo}

@@

We have two competing variants for representing individuals ---
\textit{\atTree} and \mbox{\textit{\sexprTree}} (see \ref{tree-reps}).

Individuals generated by solving inhabitation tree 
are \sexprTree{}s, which can be easily translated to \atTree{}s. \\

Before we start generating we do one important trick.
Suppose we want individuals of the type 
$\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha$
from building blocks context $\Gamma$.

All our individuals will have form $\lam{x_1\,\dots\,x_n}{M}$.
%Later in this chapter we will discuss problems occurring with variables
%when performing subtree swap in order to perform crossover.
But since those variables are same for all individuals, 
we may do following trick.

Instead of generating individuals for
$(\tau_1 \ar \dots \ar \tau_n \ar \alpha,\Gamma)$
we generate individuals for $(\alpha;\Gamma,x_1:\tau_1,\dots,x_n:\tau_n)$
and remember to add the $\lambda$-head $\lh{x_1\,\dots\,x_n}$ to
each individual when evaluating it.\\

This trick has two reasons:
\begin{enumerate}
 \item Later in this chapter we will discuss problems occurring with variables
       when performing subtree swap in order to perform crossover.
       We solve those problems by performing \textit{abstraction elimination},
       which eliminates all occurences of variables and $lambda$-abstractions
       in a \lterm. But those problems are caused by local variables. 
       Since $x_1\,\dots\,x_n$ are common for all \lterms, it is valid
       to exclude them from this elimination. Since abstraction elimination
       prolongs size of \lterms it is beneficial to lessen 
       the effect of abstraction elimination in this way.

 \item It also makes our method compatible with classical GP, where
       variables are treated as ordinary element of the set $T$
       and no $\lambda$-abstraction is used.

       This ensures that S-expressions will be generated for
       $\Gamma$ satisfying constrains of classical GP and that
       abstraction elimination will have no effect on such individuals.
\end{enumerate}

%\begin{todo}
% & několik fází, několik reprezentací
% & během generování se reprezentace mění 
%   && nejdří obsahuje lambdy i volný proměnný pak už ne
%      &&& vygeneruje se v \sexprTree
%      &&& ta se převedena na \atTree aby jsme provedli abstraction elimination
%\end{todo}

@@

In order to describe our state space to search in \lets introduce
extended version of our tree representation for typed \lterms from \ref{typed-tree-reps}.
This extension consists in adding a new kind of leaf node $(\sigma,\Gamma)$ 
standing for unfinished tree of a type $\sigma$ with context $\Gamma$ to use
as set of building blocks
\footnote{\textbf{TODO}: Tady to napsat nějak líp s tim gamma }.\\

@@

\begin{todo}
& eště se musí nějak prohrotit to, že tam je navíc ta lambda hlava
kerá ještě o něco kazí tu $depth$  - to už uplně neplatí tim že je v
uvodu popsaný to že se ta hlava ukousne ale stejně to ještě pořádně zkontrolovat
& ukázat že je to fakt to samý co kozovský klasický generování
& zavést tam pojem run pro ten jeden vnější while cyklus 
 (i když to je asi jasný z kontextu)
\end{todo}

@@

\begin{todo}
   & rozdíli při křížení ruznejch reprezentací 
      && \atTree umožňuje víc křížení než \sexprTree
          &&& každý co de v \sexprTree umí i \atTree ale obráceně to neplatí
      && \atTree je zase ekvivalentní tomu kozovskýmu  
\end{todo}
